diffEnrich: An \proglang{R} Package to Compare Functional Enrichment Between Two Experimentally-derived Groups of Genes by Connecting to the KEGG REST API

### 2019-11-15 jss-article-rnw-diffEnrich/article.Rnw
added citations for introduction

Often high throughput omics studies include a functional enrichment analysis to
glean biological insight from a list of candidate genes, proteins, metabolites,
etc. Functional enrichment examines whether the number of genes in the list
associated with a biological function or particular pathway is more than would
be expected by chance. As an example, enrichment of a particular pathway among
a list of genes that are differentially expressed after an experimental
manipulation may indicate that the pathway has been altered by that
manipulation. This analysis is rather straight forward and many solutions have
been offered (e.g., \cite{HuangDW:2009}; Kuleshov et al., 2016; Liao et al.,
2019; Subramanian et al., 2005). A wide variety of databases have also been
used to define these pathways (e.g., Kanehisa and Goto, 2000) and ontologies
(e.g., \cite{Ashburner:2000}).

@@

However, for some newer omics technologies, the background data set is hard to
define. For example, LC-MS analysis can be used to identify carbonylated
proteins ( Peterson et al., 2018; Shearn et al., 2019; Shearn et al., 2018).
With this study design, carbonylated proteins are isolated using a BH-derivation
and then LC-MS is used to identify peptides in this isolated sample. The most
appropriate background data set would be proteins present in that tissue, but
this would require a separate analytical analysis. Furthermore, most functional
enrichment analyses involve a single gene list. However, in protein modification
studies, the typical experimental design compares the presence or absence of
particular modified proteins between multiple groups.

When there are two or more gene lists to compare and the background gene list is
not clearly defined, as is often the case in protein modification experiments,
we propose a differential enrichment analysis. In this analysis, we compare the
proportion of genes/proteins from one gene list associated with a particular
pathway to the proportion of genes/proteins from a second gene list that are
associated with that pathway. To easily execute this analysis, we have designed
an R package that uses the KEGG REST API to obtain the most recent version of
the KEGG PATHWAY ( Kanehisa and Goto, 2000) database to initially identify
functional enrichment within a gene list using the entire KEGG transcriptome as
the background data set and then to identify differentially enriched pathways
between two gene lists. This R package includes a function to generate a
``differential enrichment'' graphic.

>>>>

Often high throughput omics studies include a functional enrichment analysis to
glean biological insight from a list of candidate genes, proteins, metabolites,
etc. Functional enrichment examines whether the number of genes in the list
associated with a biological function or particular pathway is more than would
be expected by chance. As an example, enrichment of a particular pathway among
a list of genes that are differentially expressed after an experimental
manipulation may indicate that the pathway has been altered by that
manipulation. This analysis is rather straight forward and many solutions have
been offered (e.g., \cite{HuangDW:2009}; \cite{Kuleshov:2016}; \cite{Liao:2019};
\cite{Subramanian:2005}). A wide variety of databases have also been
used to define these pathways (e.g., \cite{KEGG:2000}) and ontologies
(e.g., \cite{Ashburner:2000}).

@@

However, for some newer omics technologies, the background data set is hard to
define. For example, LC-MS analysis can be used to identify carbonylated
proteins ( \cite{Petersen:2018}; \cite{Shearn:2019}; \cite{Shearn:2018}).
With this study design, carbonylated proteins are isolated using a BH-derivation
and then LC-MS is used to identify peptides in this isolated sample. The most
appropriate background data set would be proteins present in that tissue, but
this would require a separate analytical analysis. Furthermore, most functional
enrichment analyses involve a single gene list. However, in protein modification
studies, the typical experimental design compares the presence or absence of
particular modified proteins between multiple groups.

When there are two or more gene lists to compare and the background gene list is
not clearly defined, as is often the case in protein modification experiments,
we propose a differential enrichment analysis. In this analysis, we compare the
proportion of genes/proteins from one gene list associated with a particular
pathway to the proportion of genes/proteins from a second gene list that are
associated with that pathway. To easily execute this analysis, we have designed
an R package that uses the KEGG REST API to obtain the most recent version of
the KEGG PATHWAY (\cite{KEGG:2000}) database to initially identify
functional enrichment within a gene list using the entire KEGG transcriptome as
the background data set and then to identify differentially enriched pathways
between two gene lists. This R package includes a function to generate a
``differential enrichment'' graphic.
